



```{r}
library(tidyverse)
library(here)

df <- read.csv(here("data", "cleaned_data.csv"))

```

```{r}
# rename cols
df<- df |> 
    dplyr::rename(no3_no2 = inorganic_nitrogen_.nitrate_and_nitrite._total,
                 tp = phosphorus_total,
                 chla = chlorophyll_a_total,
                 tss = total_suspended_solids_suspended)



```

```{r}
# concentration vs position (id)
df |> 
    ggplot(aes(x=total_distance, y=no3_no2))+
    geom_point()+
    geom_smooth(se = FALSE)+
    theme_bw()+
    facet_wrap(~site)

```

```{r}
# sample balance across position
sample_counts <- df %>%
  count(site, id, month, name = "n") %>%
  # Fill empty cells with 0 so they show up in the plot, not as gaps
  complete(site, id, month, fill = list(n = 0))

ggplot(sample_counts, aes(x = month, y = id, fill = n)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = n, color = n > max(n) / 2), size = 3, show.legend = FALSE) +
  scale_color_manual(values = c("black", "white")) +
  scale_fill_viridis_c(option = "mako", direction = -1, name = "n") +
  facet_wrap(~ site, ncol = 1) +
  labs(
    title = "Sample size by position, month, and reservoir",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  )


```

```{r}
df_long <- df %>%
  select(site, id, month, no3_no2, tp, tss, chla) %>%
  pivot_longer(
    cols      = c(no3_no2, tp, tss, chla),
    names_to  = "analyte",
    values_to = "concentration"
  ) %>%
  filter(!is.na(concentration), concentration > 0) %>%   # log scale needs > 0
  mutate(
    analyte = factor(
      analyte,
      levels = c("no3_no2", "tp", "tss", "chla"),
      labels = c("NO3+NO2", "TP", "TSS", "Chl-a")
    )
  )

ggplot(df_long, aes(x = month, y = concentration,
                    color = id, fill = id)) +
  stat_summary(fun.data = mean_cl_boot, geom = "ribbon",
               alpha = 0.18, color = NA) +
  stat_summary(fun = mean, geom = "line",
               aes(group = id), linewidth = 0.8) +
  stat_summary(fun = mean, geom = "point", size = 1.6) +
  facet_grid(analyte ~ site, scales = "free_y") +
  scale_y_log10() +
  scale_x_continuous(breaks = 1:12, labels = month.abb, minor_breaks = NULL) +
  scale_color_viridis_d(option = "viridis", end = 0.85) +
  scale_fill_viridis_d(option = "viridis", end = 0.85) +
  labs(
    title    = "Monthly nutrient concentrations by position within reservoirs",
    subtitle = "Non-parallel lines suggest position × season interaction",
    x = NULL,
    y = "Concentration (log scale)",
    color = "Position", fill = "Position"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

```

```{r}
# How many samples per year do we have?

print(df |> 
  group_by(year) |> 
  count() |> 
  ungroup(), 
  n = 30)


print(df |> 
  distinct(id, .keep_all = TRUE) |> 
  group_by(site) |> 
  count() |> 
  ungroup(), 
  n = 30)



# see hte values for the stations that only had <3 observations
obs_chk_df<- df |>
  group_by(id) %>%
  filter(n() < 3) %>%
  ungroup()


```


## Coplots

```{r}

# check if the relationship between distance from dam and no3_no2 is consistent across reservoirs
ggplot(df, aes(x = total_distance, y = no3_no2)) +
  geom_point(alpha = 0.4, size = 1.5, color = "grey40") +
  geom_smooth(se = TRUE, color = "dodgerblue4") +
  
  facet_wrap(~ site, ncol = 2) +
  
  labs(
    x = "Distance from Dam (km)",
    y = expression(NO[3] + NO[2]~"(mg/L)"),
    title = "Nitrate Concentration by Site"
  ) +
  
  theme_bw(base_size = 13) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# center hill throws the pattern a bit


