---
title: "NO3+NO2 Spatiotemporal Patterns Across Tennessee Reservoirs"
subtitle: "Hierarchical GAMM with site-specific smooths and AR(1) correlation"
author: "Spencer G. Womble"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 5,
  dpi = 150
)
```

# Overview

This analysis models nitrate plus nitrite (NO3+NO2) concentrations across three
Tennessee reservoirs (Center Hill, Dale Hollow, Percy Priest) as a function of:

- **Spatial position** within each reservoir (distance from dam),
- **Day of year** (seasonal cycle),
- **Year** (interannual trend), and
- The **interaction** between distance and day of year.

The model is a hierarchical generalized additive mixed model (GAMM) with a
global smooth plus site-specific deviation smooths, a continuous within-station
AR(1) correlation structure to handle repeated measures at fixed sampling
stations, and cyclic splines for day of year.

# Setup

```{r libraries}
library(tidyverse)   # data wrangling and plotting
library(here)        # project-relative paths
library(mgcv)        # GAMM fitting
library(gratia)      # tidy smooth plots and difference_smooths()
library(tidyr)       # crossing() for prediction grids
```

# Data import and cleaning

```{r import}
df_raw <- read.csv(here("data", "cleaned_data.csv"))
```

Two records from the 1970s sit far outside the period of consistent
monitoring and would otherwise pull the year-effect smooth. Drop them.

```{r filter-old-records}
df <- df_raw |>
  dplyr::filter(year > 1980)
```

Rename verbose source columns to compact analysis names and derive a
day-of-year column from the timestamp.

```{r rename-and-derive}
df <- df |>
  dplyr::rename(
    no3_no2 = inorganic_nitrogen_.nitrate_and_nitrite._total,
    tp      = phosphorus_total,
    chla    = chlorophyll_a_total
  ) |>
  dplyr::mutate(doy = yday(date_time))
```

# Build the modeling data frame

`id` is the sampling station, treated as a random intercept. Data are sorted
by datetime within station because the continuous AR(1) correlation in
`corCAR1()` is order-sensitive within each grouping unit.

```{r model-df}
mod_df <- df |>
  dplyr::select(
    no3_no2, total_distance, decimal_date,
    year, month, doy, date_time, id, site
  ) |>
  dplyr::mutate(
    id   = as.factor(id),
    site = factor(site, levels = c("center_hill", "dale_hollow", "percy_priest"))
  ) |>
  dplyr::arrange(date_time, id) |>
  na.omit()
```

# Model fitting

The response is log-transformed to stabilize variance and meet the assumption
of approximate Gaussian errors. The smooth structure is:

- `s(total_distance)`: global distance-from-dam effect.
- `s(total_distance, by = site, m = 1)`: site-specific deviations from the
  global distance smooth, with a first-derivative penalty so the site smooths
  shrink toward zero rather than toward a flat line.
- `s(doy, bs = "cc")`: global seasonal cycle, cyclic so Jan 1 and Dec 31 join
  smoothly.
- `s(doy, by = site, bs = "cc", m = 1)`: site-specific seasonal deviations,
  also cyclic and deviation-penalized.
- `ti(total_distance, doy, by = site, ...)`: tensor interaction allowing the
  seasonal cycle to differ between near-dam and headwater positions, with
  marginal first-derivative penalties on both axes.
- `s(year, k = 6)` and `s(year, by = site, m = 1)`: global interannual trend
  with site-specific deviations. `k = 6` is the largest basis dimension the
  model can fit without numerical issues given the year coverage.

A random intercept for `id` plus a continuous AR(1) correlation on
`decimal_date` together account for repeated measures at fixed stations.
`gamma = 1.4` follows current practice for reducing undersmoothing in mgcv.

```{r fit-gamm}
gim <- gamm(
  log(no3_no2) ~
    s(total_distance) +
    s(total_distance, by = site, m = 1) +
    s(doy, bs = "cc") +
    s(doy, by = site, bs = "cc", m = 1) +
    ti(total_distance, doy,
       by = site,
       bs = c("tp", "cc"),
       m  = list(1, 1)) +
    s(year, k = 6) +
    s(year, by = site, m = 1),
  random      = list(id = ~ 1),
  correlation = corCAR1(form = ~ decimal_date | id),
  knots       = list(doy = c(0, 366)),
  data        = mod_df,
  method      = "ML",
  gamma       = 1.4
)
```

# Inspecting fitted smooths

```{r smooth-grid, fig.height = 7}
gratia::draw(gim$gam)
```

```{r individual-smooths}
gratia::draw(gim$gam, select = "s(total_distance)")
gratia::draw(gim$gam, select = "s(doy)")
gratia::draw(gim$gam, select = "s(year)")
```

# Pairwise seasonal differences between sites

`difference_smooths()` returns the pointwise contrast between site-specific
day-of-year smooths with simultaneous confidence bands. Pass `gim$gam`
because `gamm()` returns a list and the smooth methods operate on the gam
component.

```{r doy-diffs}
doy_diffs <- gratia::difference_smooths(
  gim$gam,
  select        = "s(doy)",
  n             = 200,
  ci_level      = 0.95,
  partial_match = TRUE
)

draw(doy_diffs, ci_type = "simultaneous") +
  labs(y = expression(Delta ~ log(NO[3] ~ "+" ~ NO[2])))
```

A tidy version with a flag for intervals that exclude zero:

```{r doy-diffs-tidy}
doy_diffs_df <- doy_diffs |>
  dplyr::mutate(
    sig  = (.lower_ci > 0) | (.upper_ci < 0),
    pair = paste(.level_1, "vs", .level_2)
  )
```

## Interpretation

**Center Hill vs Dale Hollow.** Strong seasonal contrast. Center Hill runs
about 1 log-unit (roughly 2.7x) higher in winter and late fall, converges
with Dale Hollow through summer (the ribbon spans zero roughly from doy 150
to 225), and diverges again into fall. Warm-season concentrations are
similar at the two reservoirs; cool-season concentrations are not.

**Center Hill vs Percy Priest.** Effectively indistinguishable across the
observed range. The pointwise interval includes zero throughout, so no
seasonal difference between these two reservoirs is supported by this fit.

**Dale Hollow vs Percy Priest.** The most pronounced contrast. Dale Hollow
is roughly 1.3 log-units lower (about 3.7x lower) in spring and late fall,
then crosses to about 0.7 higher (about 2x) around doy 175 to 200. This is
a phase inversion of the seasonal pattern rather than a simple amplitude
shift.

Taken together, Center Hill and Percy Priest behave similarly through the
year and Dale Hollow is the outlier, with much lower cool-season
concentrations and a relatively elevated summer signal. That is consistent
with Dale Hollow's well-known oligotrophy (low background loading,
clear-water Obey River basin) and points toward a discussion of mixing
regime, stratification timing, and watershed inputs versus the Caney Fork
and Stones River basins.

# Distance-by-DOY interaction visualization

Distance scenarios are defined within each site because the three reservoirs
differ in length. The 10th, 50th, and 90th percentiles of `total_distance`
within each site define near-dam, mid-reservoir, and headwater positions.

```{r ti-grid}
scenarios <- mod_df |>
  dplyr::group_by(site) |>
  dplyr::reframe(
    position       = factor(
      c("Near-dam", "Mid-reservoir", "Headwater"),
      levels = c("Near-dam", "Mid-reservoir", "Headwater")
    ),
    total_distance = quantile(total_distance, c(0.10, 0.50, 0.90))
  )

grid <- scenarios |>
  tidyr::crossing(doy = seq(1, 366, length.out = 200)) |>
  dplyr::mutate(year = mean(mod_df$year))
```

Predictions are taken from the GAM component; the random effect lives on
`$lme` and is not needed for population-level predictions.

```{r ti-predict}
preds <- predict(gim$gam, newdata = grid, se.fit = TRUE)

grid <- grid |>
  dplyr::mutate(
    fit_log = preds$fit,
    se_log  = preds$se.fit,
    fit     = exp(fit_log),
    lwr     = exp(fit_log - 1.96 * se_log),
    upr     = exp(fit_log + 1.96 * se_log)
  )
```

```{r ti-plot, fig.height = 6.5}
ggplot(grid, aes(doy, fit, color = position, fill = position)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8) +
  facet_grid(position ~ site) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  scale_y_log10() +
  labs(
    title    = "Seasonal NO3+NO2 by reservoir position",
    subtitle = "Position defined by within-site 0.10, 0.50, and 0.90 quantiles of distance from dam (log10 y-axis)",
    x        = "Day of year",
    y        = expression("Predicted NO"[3] * "+" * "NO"[2] * " (mg/L)"),
    color    = "Position",
    fill     = "Position"
  ) +
  theme_bw() +
  theme(text = element_text(size = 14))
```

# Pairwise interannual differences between sites

```{r year-diffs}
year_diffs <- gratia::difference_smooths(
  gim$gam,
  select        = "s(year)",
  n             = 200,
  ci_level      = 0.95,
  partial_match = TRUE
)

draw(year_diffs, ci_type = "simultaneous") +
  labs(y = expression(Delta ~ log(NO[3] ~ "+" ~ NO[2])))
```
